A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2024-01-15 paper.Rmd
sec 1-2 update

The lineup protocol places a data plot in a field of null plots, allowing for a comparison of patterns due purely by chance to what is perceived in the data plot. For residual analysis this is especially helpful for gauging whether there is \emph{no} pattern. (Figure \ref{fig:first-example-lineup} shows an example of a lineup of residual plots.) In its strict use, one would insist that the data plot is not seen before seeing the lineup, so that the observer doesn't know which is the true plot. When used this way, it provides an objective test for data plots. \cite{majumdervalidation2013} validated that results from lineups assessed by human observers performed similarly to conventional tests. One wouldn't use a lineup when a conventional test exists and is adequate because it is more manually expensive to conduct. However, where no adequate conventional test exists, it is invaluable, as shown by \cite{loy2013diagnostic}. Here we use the lineup as a vehicle to rigorously explore why experts advise that residual plots are indispensable despite the prevalence of numerical tests.

@@

Graphical summaries where residuals are plotted against fitted values, or other functions of the predictors (expected to be approximately orthogonal to the residuals) are considered to be the most important residual plots by @cook1999applied. Figure \ref{fig:residual-plot-common-departures}A shows an example of an ideal residual plot where points are symmetrically distributed around the horizontal zero line (red), with no discernible patterns. There can be various types of departures from this ideal pattern. Non-linearity, heteroskedasticity and non-normality, shown in Figures \ref{fig:residual-plot-common-departures}B, \ref{fig:residual-plot-common-departures}C, and \ref{fig:residual-plot-common-departures}D are three commonly checked departures.

@@

Model misspecification occurs if functions of predictors that needed to accurately describe the relationship with the response are incorrectly specified. This includes instances where a higher-order polynomial term of a predictor is wrongfully omitted. Any non-linear pattern visible in the residual plot could be indicative of this problem. An example residual plot containing visual pattern of non-linearity is shown in Figure \ref{fig:residual-plot-common-departures}B. One can clearly observe the "S-shape" from the residual plot, which corresponds to cubic term that should have been included in the model.

@@

Figure \ref{fig:residual-plot-common-departures}D shows a scatterplot where the residuals have a skewed distribution, as seen by the uneven vertical spread. Unlike non-linearity and heteroskedasticity, non-normality is usually detected with a different type of residual plot: a histogram or a normal probability plot.  Because we focus on scatterplots, non-normality is not one of the departures examined in depth in this paper. (@loy2016variations discuss related work on non-normality checking.)

@@

With large sample sizes, hypothesis tests may reject the null hypothesis when there is only a small effect. (A good discussion can be found in @kirk1996practical.) While such rejections may be statistically correct, their sensitivity may render the results impractical. A key goal of residual plot diagnostics is to identify potential issues that could lead to incorrect conclusions or errors in subsequent analyses, but minor defects in the model are unlikely to have a significant impact and may be best disregarded for practical purposes. The experiment discussed in this paper specifically addresses this tension between statistical and practical statistics.

@@

The examination of data plots to infer signal or patterns (or lack thereof) is fraught with variation in the human ability to interpret and decode the information embedded in a graph [@cleveland1984graphical]. Human examination of diagnostic plots can feel subjective. Ideally we would expect visual evaluation to be done objectively with low human-to-human variation in reading data plots, but we cannot assume this is the case.

In practice, over-interpretation of a single plot is common.  For instance, @roy2015using described a published example where authors over-interpreted separation between gene groups from a two-dimensional projection of a linear discriminant analysis even when there were no differences in the expression levels between the gene groups. One solution to over-interpretation is to examine the plot in the context of natural sampling variability assumed by the model, called the lineup protocol, as proposed in @buja2009statistical.  @majumder2013validation showed that the lineup protocol is analogous to the null hypothesis significance testing framework. The protocol consists of $m$ randomly placed plots, where one plot is the data plot, and the remaining $m - 1$ plots, referred to as the _null plots_, are constructed using the same graphical procedure as the data plot but the data is replaced with null data that is generated in a manner consistent with the null hypothesis, $H_0$. Then, an observer who has not seen the data plot is asked to point out the most different plot from the lineup. Under $H_0$, it is expected that the data plot would have no distinguishable difference from the null plots, and the probability that the observer correctly picks the data plot is $1/m$. If one rejects $H_0$ as the observer correctly picks the data plot, then the Type I error of this test is $1/m$. This protocol requires a priori specification of $H_0$ (or at least a null data generating mechanism), much like the requirement of knowing the sampling distribution of the test statistic in null hypothesis significance testing framework.

@@

The effectiveness of lineup protocol for regression analysis has been validated by @majumder2013validation under relatively simple settings with up to two predictors. Their results suggest that visual tests are capable of testing the significance of a single predictor with a similar power to a t-test, though they express that in general it is unnecessary to use visual inference if there exists a corresponding conventional test, and they do not expect the visual test to perform equally well as the conventional test. In their third experiment, where the contamination of the data violate the assumptions of the convetional test, visual test outperforms the conventional test by a large margin. This is encouraging, as it supports the use of visual inference in situations where there are no existing numerical testing procedures. Visual inference has also been integrated into diagnostics for hierarchical linear models where the lineup protocol is used to judge the assumptions of linearity, normality and constant error variance for both the level-1 and level-2 residuals \citep{loy2013diagnostic, loy2014hlmdiag, loy2015you}.

@@

The supplementary material is available at https://github.com/TengMCing/lineup_residual_diagnostics/blob/master/appendix.pdf. It includes more details about the experimental setup, the derivation of the effect size, the effect of data collection period, and the estimate of $\alpha$.

The R package `visage` is currently available at https://github.com/TengMCing/visage.

>>>>

The lineup protocol places a data plot in a field of null plots, allowing for a comparison of patterns due purely by chance to what is perceived in the data plot. For residual analysis this is especially helpful for gauging whether there is \emph{no} pattern. (Figure \ref{fig:first-example-lineup} shows an example of a lineup of residual plots.) In its strict use, one would insist that the data plot is not seen before seeing the lineup, so that the observer doesn't know which is the true plot. When used this way, it provides an objective test for data plots. \cite{majumdervalidation2013} validated that results from lineups assessed by human observers performed similarly to conventional tests. One would not use a lineup when a conventional test exists and is adequate because it is more manually expensive to conduct. However, where no adequate conventional test exists, it is invaluable, as shown by \cite{loy2013diagnostic}. Here we use the lineup as a vehicle to rigorously explore why experts advise that residual plots are indispensable despite the prevalence of numerical tests.

@@

Graphical summaries where residuals are plotted against fitted values, or other functions of the predictors (expected to be approximately orthogonal to the residuals) are considered to be the most important residual plots by @cook1999applied. Figure \ref{fig:residual-plot-common-departures}A shows an example of an ideal residual plot where points are symmetrically distributed around the horizontal zero line (red), with no discernible patterns. There can be various types of departures from this ideal pattern. Non-linearity, heteroskedasticity and non-normality, shown in Figures \ref{fig:residual-plot-common-departures}B, \ref{fig:residual-plot-common-departures}C, and \ref{fig:residual-plot-common-departures}D, respectively, are three commonly checked departures.

@@

Model misspecification occurs if functions of predictors that needed to accurately describe the relationship with the response are incorrectly specified. This includes instances where a higher-order polynomial term of a predictor is wrongfully omitted. Any non-linear pattern visible in the residual plot could be indicative of this problem. An example residual plot containing visual pattern of non-linearity is shown in Figure \ref{fig:residual-plot-common-departures}B. One can clearly observe the "S-shape" from the residual plot, which corresponds to the cubic term that should have been included in the model.

@@

Figure \ref{fig:residual-plot-common-departures}D shows a scatterplot where the residuals have a skewed distribution, as seen by the uneven vertical spread. Unlike non-linearity and heteroskedasticity, non-normality is usually detected with a different type of residual plot: a histogram or a normal probability plot.  Because we focus on scatterplots, non-normality is not one of the departures examined in depth in this paper. [@loy2016variations discuss related work on non-normality checking.]

@@

With large sample sizes, hypothesis tests may reject the null hypothesis when there is only a small effect. (A good discussion can be found in @kirk1996practical.) While such rejections may be statistically correct, their sensitivity may render the results impractical. A key goal of residual plot diagnostics is to identify potential issues that could lead to incorrect conclusions or errors in subsequent analyses, but minor defects in the model are unlikely to have a significant impact and may be best disregarded for practical purposes. The experiment discussed in this paper specifically addresses this tension between statistical and practical significance.

@@

The examination of data plots to infer signals or patterns (or lack thereof) is fraught with variation in the human ability to interpret and decode the information embedded in a graph [@cleveland1984graphical]. In practice, over-interpretation of a single plot is common.  For instance, @roy2015using described a published example where authors over-interpreted separation between gene groups from a two-dimensional projection of a linear discriminant analysis even when there were no differences in the expression levels between the gene groups.

One solution to over-interpretation is to examine the plot in the context of natural sampling variability assumed by the model, called the lineup protocol, as proposed in @buja2009statistical.  @majumder2013validation showed that the lineup protocol is analogous to the null hypothesis significance testing framework. The protocol consists of $m$ randomly placed plots, where one plot is the data plot, and the remaining $m - 1$ plots, referred to as the _null plots_, are constructed using the same graphical procedure as the data plot but the data is replaced with null data that is generated in a manner consistent with the null hypothesis, $H_0$. Then, an observer who has not seen the data plot is asked to point out the most different plot from the lineup. Under $H_0$, it is expected that the data plot would have no distinguishable difference from the null plots, and the probability that the observer correctly picks the data plot is $1/m$. If one rejects $H_0$ as the observer correctly picks the data plot, then the Type I error of this test is $1/m$. This protocol requires a priori specification of $H_0$ (or at least a null data generating mechanism), much like the requirement of knowing the sampling distribution of the test statistic in null hypothesis significance testing framework.

@@

The effectiveness of lineup protocol for regression analysis has been validated by @majumder2013validation under relatively simple settings with up to two predictors. Their results suggest that visual tests are capable of testing the significance of a single predictor with a similar power to a t-test, though they express that in general it is unnecessary to use visual inference if there exists a corresponding conventional test, and they do not expect the visual test to perform equally well as the conventional test. In their third experiment, where the contamination of the data violate the assumptions of the conventional test, visual test outperforms the conventional test by a large margin. This supports the use of visual inference in situations where there are no existing numerical testing procedures. Visual inference has also been integrated into diagnostics for hierarchical linear models where the lineup protocol is used to judge the assumptions of linearity, normality and constant error variance for both the level-1 and level-2 residuals \citep{loy2013diagnostic, loy2014hlmdiag, loy2015you}.

@@

The supplementary material is available at https://github.com/TengMCing/lineup_residual_diagnostics/blob/master/appendix.pdf. It includes more details about the experimental setup, the derivation of the effect size, the effect of data collection period, and the estimate of $\alpha$. The R package `visage` is currently available at https://github.com/TengMCing/visage.

### 2024-01-15 paper.tex
sec 1-2 update

Graphical summaries where residuals are plotted against fitted values,
or other functions of the predictors (expected to be approximately
orthogonal to the residuals) are considered to be the most important
residual plots by \citet{cook1999applied}. Figure
\ref{fig:residual-plot-common-departures}A shows an example of an ideal
residual plot where points are symmetrically distributed around the
horizontal zero line (red), with no discernible patterns. There can be
various types of departures from this ideal pattern. Non-linearity,
heteroskedasticity and non-normality, shown in Figures
\ref{fig:residual-plot-common-departures}B,
\ref{fig:residual-plot-common-departures}C, and
\ref{fig:residual-plot-common-departures}D are three commonly checked
departures.

@@

Model misspecification occurs if functions of predictors that needed to
accurately describe the relationship with the response are incorrectly
specified. This includes instances where a higher-order polynomial term
of a predictor is wrongfully omitted. Any non-linear pattern visible in
the residual plot could be indicative of this problem. An example
residual plot containing visual pattern of non-linearity is shown in
Figure \ref{fig:residual-plot-common-departures}B. One can clearly
observe the ``S-shape'' from the residual plot, which corresponds to
cubic term that should have been included in the model.

@@

Figure \ref{fig:residual-plot-common-departures}D shows a scatterplot
where the residuals have a skewed distribution, as seen by the uneven
vertical spread. Unlike non-linearity and heteroskedasticity,
non-normality is usually detected with a different type of residual
plot: a histogram or a normal probability plot. Because we focus on
scatterplots, non-normality is not one of the departures examined in
depth in this paper. (\citet{loy2016variations} discuss related work on
non-normality checking.)

@@

With large sample sizes, hypothesis tests may reject the null hypothesis
when there is only a small effect. (A good discussion can be found in
\citet{kirk1996practical}.) While such rejections may be statistically
correct, their sensitivity may render the results impractical. A key
goal of residual plot diagnostics is to identify potential issues that
could lead to incorrect conclusions or errors in subsequent analyses,
but minor defects in the model are unlikely to have a significant impact
and may be best disregarded for practical purposes. The experiment
discussed in this paper specifically addresses this tension between
statistical and practical statistics.

@@

The examination of data plots to infer signal or patterns (or lack
thereof) is fraught with variation in the human ability to interpret and
decode the information embedded in a graph
\citep{cleveland1984graphical}. Human examination of diagnostic plots
can feel subjective. Ideally we would expect visual evaluation to be
done objectively with low human-to-human variation in reading data
plots, but we cannot assume this is the case.

In practice, over-interpretation of a single plot is common. For
instance, \citet{roy2015using} described a published example where
authors over-interpreted separation between gene groups from a
two-dimensional projection of a linear discriminant analysis even when
there were no differences in the expression levels between the gene
groups. One solution to over-interpretation is to examine the plot in
the context of natural sampling variability assumed by the model, called
the lineup protocol, as proposed in \citet{buja2009statistical}.
\citet{majumder2013validation} showed that the lineup protocol is
analogous to the null hypothesis significance testing framework. The
protocol consists of \(m\) randomly placed plots, where one plot is the
data plot, and the remaining \(m - 1\) plots, referred to as the
\emph{null plots}, are constructed using the same graphical procedure as
the data plot but the data is replaced with null data that is generated
in a manner consistent with the null hypothesis, \(H_0\). Then, an
observer who has not seen the data plot is asked to point out the most
different plot from the lineup. Under \(H_0\), it is expected that the
data plot would have no distinguishable difference from the null plots,
and the probability that the observer correctly picks the data plot is
\(1/m\). If one rejects \(H_0\) as the observer correctly picks the data
plot, then the Type I error of this test is \(1/m\). This protocol
requires a priori specification of \(H_0\) (or at least a null data
generating mechanism), much like the requirement of knowing the sampling
distribution of the test statistic in null hypothesis significance
testing framework.

@@

The effectiveness of lineup protocol for regression analysis has been
validated by \citet{majumder2013validation} under relatively simple
settings with up to two predictors. Their results suggest that visual
tests are capable of testing the significance of a single predictor with
a similar power to a t-test, though they express that in general it is
unnecessary to use visual inference if there exists a corresponding
conventional test, and they do not expect the visual test to perform
equally well as the conventional test. In their third experiment, where
the contamination of the data violate the assumptions of the convetional
test, visual test outperforms the conventional test by a large margin.
This is encouraging, as it supports the use of visual inference in
situations where there are no existing numerical testing procedures.
Visual inference has also been integrated into diagnostics for
hierarchical linear models where the lineup protocol is used to judge
the assumptions of linearity, normality and constant error variance for
both the level-1 and level-2 residuals
\citep{loy2013diagnostic, loy2014hlmdiag, loy2015you}.

@@

The supplementary material is available at
\url{https://github.com/TengMCing/lineup_residual_diagnostics/blob/master/appendix.pdf}.
It includes more details about the experimental setup, the derivation of
the effect size, the effect of data collection period, and the estimate
of \(\alpha\).

The R package \texttt{visage} is currently available at
\url{https://github.com/TengMCing/visage}.

>>>>

Graphical summaries where residuals are plotted against fitted values,
or other functions of the predictors (expected to be approximately
orthogonal to the residuals) are considered to be the most important
residual plots by \citet{cook1999applied}. Figure
\ref{fig:residual-plot-common-departures}A shows an example of an ideal
residual plot where points are symmetrically distributed around the
horizontal zero line (red), with no discernible patterns. There can be
various types of departures from this ideal pattern. Non-linearity,
heteroskedasticity and non-normality, shown in Figures
\ref{fig:residual-plot-common-departures}B,
\ref{fig:residual-plot-common-departures}C, and
\ref{fig:residual-plot-common-departures}D, respectively, are three
commonly checked departures.

@@

Model misspecification occurs if functions of predictors that needed to
accurately describe the relationship with the response are incorrectly
specified. This includes instances where a higher-order polynomial term
of a predictor is wrongfully omitted. Any non-linear pattern visible in
the residual plot could be indicative of this problem. An example
residual plot containing visual pattern of non-linearity is shown in
Figure \ref{fig:residual-plot-common-departures}B. One can clearly
observe the ``S-shape'' from the residual plot, which corresponds to the
cubic term that should have been included in the model.

@@

Figure \ref{fig:residual-plot-common-departures}D shows a scatterplot
where the residuals have a skewed distribution, as seen by the uneven
vertical spread. Unlike non-linearity and heteroskedasticity,
non-normality is usually detected with a different type of residual
plot: a histogram or a normal probability plot. Because we focus on
scatterplots, non-normality is not one of the departures examined in
depth in this paper. \citep[ discuss related work on non-normality
checking.]{loy2016variations}

@@

With large sample sizes, hypothesis tests may reject the null hypothesis
when there is only a small effect. (A good discussion can be found in
\citet{kirk1996practical}.) While such rejections may be statistically
correct, their sensitivity may render the results impractical. A key
goal of residual plot diagnostics is to identify potential issues that
could lead to incorrect conclusions or errors in subsequent analyses,
but minor defects in the model are unlikely to have a significant impact
and may be best disregarded for practical purposes. The experiment
discussed in this paper specifically addresses this tension between
statistical and practical significance.

@@

The examination of data plots to infer signals or patterns (or lack
thereof) is fraught with variation in the human ability to interpret and
decode the information embedded in a graph
\citep{cleveland1984graphical}. In practice, over-interpretation of a
single plot is common. For instance, \citet{roy2015using} described a
published example where authors over-interpreted separation between gene
groups from a two-dimensional projection of a linear discriminant
analysis even when there were no differences in the expression levels
between the gene groups.

One solution to over-interpretation is to examine the plot in the
context of natural sampling variability assumed by the model, called the
lineup protocol, as proposed in \citet{buja2009statistical}.
\citet{majumder2013validation} showed that the lineup protocol is
analogous to the null hypothesis significance testing framework. The
protocol consists of \(m\) randomly placed plots, where one plot is the
data plot, and the remaining \(m - 1\) plots, referred to as the
\emph{null plots}, are constructed using the same graphical procedure as
the data plot but the data is replaced with null data that is generated
in a manner consistent with the null hypothesis, \(H_0\). Then, an
observer who has not seen the data plot is asked to point out the most
different plot from the lineup. Under \(H_0\), it is expected that the
data plot would have no distinguishable difference from the null plots,
and the probability that the observer correctly picks the data plot is
\(1/m\). If one rejects \(H_0\) as the observer correctly picks the data
plot, then the Type I error of this test is \(1/m\). This protocol
requires a priori specification of \(H_0\) (or at least a null data
generating mechanism), much like the requirement of knowing the sampling
distribution of the test statistic in null hypothesis significance
testing framework.

@@

The effectiveness of lineup protocol for regression analysis has been
validated by \citet{majumder2013validation} under relatively simple
settings with up to two predictors. Their results suggest that visual
tests are capable of testing the significance of a single predictor with
a similar power to a t-test, though they express that in general it is
unnecessary to use visual inference if there exists a corresponding
conventional test, and they do not expect the visual test to perform
equally well as the conventional test. In their third experiment, where
the contamination of the data violate the assumptions of the
conventional test, visual test outperforms the conventional test by a
large margin. This supports the use of visual inference in situations
where there are no existing numerical testing procedures. Visual
inference has also been integrated into diagnostics for hierarchical
linear models where the lineup protocol is used to judge the assumptions
of linearity, normality and constant error variance for both the level-1
and level-2 residuals
\citep{loy2013diagnostic, loy2014hlmdiag, loy2015you}.

@@

The supplementary material is available at
\url{https://github.com/TengMCing/lineup_residual_diagnostics/blob/master/appendix.pdf}.
It includes more details about the experimental setup, the derivation of
the effect size, the effect of data collection period, and the estimate
of \(\alpha\). The R package \texttt{visage} is currently available at
\url{https://github.com/TengMCing/visage}.
